\vspace{-1cm} Model linearity breeds contempt: using non-linear models to uncover general biogeographical patterns

### 2021-03-02 manuscript/manuscript.tex
Manuscript changes and data restructuring

%Title paper
\title{\vspace{-1cm}
Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions}
% Model familiarity breeds contempt: simple models to compare properties of species' distributions
% Model simplicity breeds contempt: simple models to compare properties of species' distributions
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

@@

One of the central goals of ecology is to understand the ways species are distributed across space and time (ref). Over the last two decades, ecologists have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000}. These models estimate species' realized niches using several covariates, including environmental variables \citep{Guisan}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}. More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species \citep{ovaskainenHowMakeMore2017}. The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time. However, while these factors can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019}, the interpretation of the corresponding parameter estimates has been often questioned \citep{gotelliEmpiricalBayesApproach2010, harrisInferringSpeciesInteractions2016, thurmanTestingLinkSpecies2019}. This was best illustrated by \citet{blanchetCooccurrenceNotEvidence2020}, who used basic statistical arguments to highlight the artefactual nature of the link between co-occurrence and species' ecological interactions drawn by some distribution models.
% Other references for biotic interaction=bad: dormannBioticInteractionsSpecies2018

@@

\subsubsection*{Distribution data}
We used data describing the distribution of 798 species across 912 sites covering most of the mountain region of the Western Alps in the Canton de Vaud (Switzerland; \citealt{scherrerEcologicalIndicatorValues2019}). Each of these sites is a $8\times 8\,\text{m}$ plot placed somewhere along an elevation range from $375\,\text{m}$ to $3210\,\text{m}$. In all sites, presence/absence data as well as Braun-Blanquet abundance-dominance classes were recorded for all species. Additionally, we used meteorological data provided by \citet{scherrerEcologicalIndicatorValues2019}, containing multiple variables characterizing the climate in each site at high spatial resolution ($25\,\text{m}$). This dataset was compiled based on 30 years (1961–1990) of records from national weather stations. Since most of the data is highly correlated, we calculated the main axes of variation of the following variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (see Supplementary Methods; Supplementary Fig. 1). %Missing TabsY

@@

\subsubsection*{Model priors}
The model structure described above allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions contained in the floristic database. To do so, we define the prior distributions for the parameters in model (\ref{eq:baseline-response}) as:
\begin{equation} 
\begin{split}
\beta_{i}  & \sim \text{MVNormal}\left(\hat{\beta}, \Sigma^{\beta}\right)\\
\text{log}(\gamma_{i})  & \sim \text{MVNormal}\left(\hat{\gamma}, \Sigma^{\gamma}\right)\\
\text{log}(\alpha_{i}) & \sim \text{Normal}\left(\hat{\alpha}, \sigma_{\alpha}\right)\\
\hat{\beta}, 
\hat{\gamma}, 
\hat{\alpha}  & \sim \text{Normal}\left(0,1\right)\\
\sigma_{\alpha}  & \sim \text{Exponential}\left(1\right)
\end{split}
\label{eq:baseline-priors}
\end{equation} 
where parameters $\gamma_i$ and $\beta_i$ are expressed as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta}$ and $\Sigma^{\gamma}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively. We define these variance-covariance matrices as follows:
\begin{equation} 
\Sigma_{ij} = \eta\,\text{exp}\left(-\rho {D_{ij}}^2\right) + \delta_{ij} \sigma ,
\label{eq:covariance-baseline}
\end{equation}
where $\Sigma_{ij}$ characterizes the covariance between any pair of species $i$ and $j$. Notice that such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information. In the floristic database, this information is represented by the set of ordinal specified for the different species. While there are many different ways to turn ordinal data into distance matrices, we choose to use a mixed-membership stochastic block model because it allows us to deal with cases of missing data (see Supplementary Methods for extended details; \citealt{godoy-loriteAccurateScalableSocial2016}). In each covariance matrix, the hyperparameter $\rho$ determines the rate of decline of the covariance between any two species, and $\eta$ defines its maximum value. The hyperparameter $\sigma$ describes the additional covariance between the different observations for any given species. For all these hyperparameters, we choose weekly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$.

@@

\subsubsection*{Sampling the posterior}
We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}. Sampling models like the ones described above can be computationally very expensive. This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}). Therefore, we focus on those species for which we have more than 30 occurrences when modelling ordinal data, which is the case for the majority of the results of this work. When using presence/absence data, we limit our study to those species for which have more than 10 occurrences.

@@

One way to test the extend to which species' physiological traits inform the skewness of the distribution of any species $i$ is by characterizing $\lambda_{i}$ as a Gaussian process. That is, we can consider the prior distribution for $\lambda_{i}$ as a multivariate normal with a variance covariance matrix $\Sigma^{\lambda}$. As described before, this variance-covariance matrix can be estimated by the model following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are.

@@

\section*{Results}
% We first use the baseline model to understand how species are distributed along environmental gradient. As expected alpha declines with altitude, implying that the overall...  study the relationship between mean and variance. We found positive relationship... Hypothesis! Similar results were found using binary data. Different species show different correlaitons, invasive present less prevasive... When adding additional covariate, such the second axis of our PCA, this relationship vanishes... (see Supplementary)

%Other interesting comparisons... The value of adding plant physiology as prior information for both the mean and standard deviation of the distributions. Informative in both case, specially for... Capturing some of bla bla bla.

@@

% Next we studied the skewness of distributions, outpreforming the other models. When studying the parameter, we found that the mean skewness of species distributions declined... Hypothesis! When accounting for plant physiology as priors for the skewness of distribution, we found this to be ... Figure and Hypothesis. The model also showcased the Rapopor's hypothesis

>>>>

%Title paper
\title{\vspace{-1cm}
Model linearity breeds contempt: using non-linear models to uncover general biogeographical patterns}
% Model familiarity breeds contempt: simple models to compare properties of species' distributions
% Model simplicity breeds contempt: simple models to compare properties of species' distributions
% Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

@@

One of the central goals of ecology is to understand the ways species are distributed across space and time (ref). Over the last two decades, ecologists have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000}. These models estimate species' realized niches using several covariates, including environmental variables \citep{Guisan}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}. More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species \citep{ovaskainenHowMakeMore2017}. The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time. However, while these factors can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019}, the interpretation of the corresponding parameter estimates has been recently questioned \citep{harrisInferringSpeciesInteractions2016, thurmanTestingLinkSpecies2019, poggiatoInterpretationsJointModeling2021}. This was best illustrated by \citet{blanchetCooccurrenceNotEvidence2020}, who used basic statistical arguments to highlight the artefactual nature of the link between co-occurrence and species' ecological interactions drawn by some distribution models.
% Other references for biotic interaction=bad: dormannBioticInteractionsSpecies2018

@@

\subsubsection*{Distribution data}
We used data describing the distribution of 798 species across 912 sites covering most of the mountain region of the Western Alps in the Canton de Vaud (Switzerland; \citealt{scherrerEcologicalIndicatorValues2019}). Each of these sites is a $8\times 8\,\text{m}$ plot placed somewhere along an elevation range from $375\,\text{m}$ to $3210\,\text{m}$. In all sites, presence/absence data as well as Braun-Blanquet abundance-dominance classes were recorded for all species. Additionally, we used meteorological data provided by \citet{scherrerEcologicalIndicatorValues2019}, containing multiple variables characterizing the climate in each site at high spatial resolution ($25\,\text{m}$). This dataset was compiled based on 30 years (1961–1990) of records from national weather stations. Since most of the data is highly correlated, we calculated the main axes of variation of the following scaled variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (see Supplementary Methods; Supplementary Fig. 1). %Missing TabsY

@@

\subsubsection*{Model priors}
The model structure described above allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions contained in the floristic database. To do so, we define the prior distributions for the parameters in model (\ref{eq:baseline-response}) as:
\begin{equation} 
\begin{split}
\beta_{i}  & \sim \text{MVNormal}\left(\hat{\beta}, \Sigma^{\beta}\right)\\
\text{log}(\gamma_{i})  & \sim \text{MVNormal}\left(\hat{\gamma}, \Sigma^{\gamma}\right)\\
\text{log}(\alpha_{i}) & \sim \text{Normal}\left(\hat{\alpha}, \sigma_{\alpha}\right)\\
\hat{\beta}, 
\hat{\gamma}, 
\hat{\alpha}  & \sim \text{Normal}\left(0,1\right)\\
\sigma_{\alpha}  & \sim \text{Exponential}\left(1\right)
\end{split}
\label{eq:baseline-priors}
\end{equation} 
where parameters $\gamma_i$ and $\beta_i$ are expressed as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta}$ and $\Sigma^{\gamma}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively. We define these variance-covariance matrices as follows:
\begin{equation} 
\Sigma_{ij} = \eta\,\text{exp}\left(-\rho {D_{ij}}^2\right) + \delta_{ij} \sigma ,
\label{eq:covariance-baseline}
\end{equation}
where $\Sigma_{ij}$ characterizes the covariance between any pair of species $i$ and $j$, and $\delta_{ij}$ is the Kronecker delta. Notice that such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information. In the floristic database, this information is represented by the set of ordinal specified for the different species. While there are many different ways to turn ordinal data into distance matrices, we choose to use a mixed-membership stochastic block model because it allows us to deal with cases of missing data (see Supplementary Methods for extended details; \citealt{godoy-loriteAccurateScalableSocial2016}). In each covariance matrix, the hyperparameter $\rho$ determines the rate of decline of the covariance between any two species, and $\eta$ defines its maximum value. The hyperparameter $\sigma$ describes the additional covariance between the different observations for any given species. For all these hyperparameters, we choose weekly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$.

@@

\subsubsection*{Sampling the posterior}
We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}. Sampling models like the ones described above can be computationally very expensive. This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}). Therefore, we focus on those species for which we have more than 10 occurrences when modelling binary data, which is the case for the majority of the results of this work. When using ordinal data, we limit our study to those species for which have more than 50 occurrences.

@@

One way to test the extend to which species' physiological traits inform the skewness of the distribution of any species $i$ is by characterizing $\lambda_{i}$ as a Gaussian process. That is, we can consider the prior distribution for $\lambda_{i}$ as a multivariate normal with a variance covariance matrix $\Sigma^{\lambda}$. As described before, this variance-covariance matrix can be estimated by the model following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are. Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.

@@

\section*{Results}
We studied the distribution data to characterize species' realized niches. To do so, we selected the main axis of variation of all environmental variables (Supplementary Fig.~XX). Then, using the presence or absence of species per site as response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation. Studying the relationship between the mean and variance of the distributions, we found clear evidence of the Rapopor's rule. Specifically, we found that species' range sizes decline with elevation and temperature (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model are negatively correlated; Fig.~\ref{fig:correlation}). The relationship was also found when using instead ordinal data (Supplementary Fig.~XX), and it does not seem to depend on species' type or reflect species' abundance change tendency over the years (Supplementary Fig.~XX). This pattern was not present along the second axis of variation of the environmental variables (Supplementary Fig.~XX). The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships. In particular, we found the amplitude of distributions to be positively and negatively correlated with the mean and the variance along the elevation gradient, respectively (Supplementary Fig.~XX).

Maintaining the symmetry of species' distributions, we then allowed the kurtosis---or shape of the tails---of these to vary in different ways. To do so, we changed the response curve of our Bayesian model to follow either a generalized error distribution (Eq.~\ref{eq:generalized-response}) or a non-standardized Student's t-distribution (Eq.~\ref{eq:student-response}).
%The posterior distributions of the hyperparameters in Eq.~(\ref{eq:covariance-baseline}) revealed the value of this prior information on the different parameter estimates.  to understand the value of our prior information.

We then studied the level of information contained in physiological traits to explain the properties of distributions. Specifically, using Eq.~(\ref{eq:covariance-complex}), we accounted for both physiological and environmental traits in the variance-covariance structure for the mean and standard deviation of distributions. We information 
As expected, the posterior distributions also showcase a positive relationship between... This is expected as ... Likewise, studying ... Following this, we added additional ... using Eq.~(\ref{eq:covariance-complex}). When studying plant... 
%Other interesting comparisons... The value of adding plant physiology as prior information for both the mean and standard deviation of the distributions. Informative in both case, specially for... Capturing some of bla bla bla.

@@

% Next we studied the skewness of distributions, outpreforming the other models. When studying the parameter, we found that the mean skewness of species distributions declined... Hypothesis! When accounting for plant physiology as priors for the skewness of distribution, we found this to be ... Figure and Hypothesis. The model also showcased the Rapopor's hypothesis...
